---
title: "`ciftiTools` Demo"
author: "Damon Pham & Amanda Mejia"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
    keep_md: true
    mathjax: null
vignette: >
  %\VignetteIndexEntry{`ciftiTools` Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
library(knitr)
knitr::opts_chunk$set(autodep = TRUE, cache = FALSE)
```

`ciftiTools` is an R package for working with CIFTI-2 format brain imaging data. Used in conjunction with GIFTI surface geometry files, CIFTI files enable surface-based analysis of gray matter data, which has several advantages over traditional volumetric/voxel-based analysis. Because of this, the CIFTI-2 format is used by recent neuroimaging studies including the Human Connectome Project (HCP). `ciftiTools` supports reading, writing, visualizing, resampling, and other operations for CIFTI files with the `".dscalar.nii"`, `".dtseries.nii"`, and `".dlabel.nii"` intents. Several of these operations are made possible by the [Connectome Workbench](https://www.humanconnectome.org/software/connectome-workbench). 

To get started, the first time you use `ciftiTools`, install it from either CRAN with `install.packages()` or Github with `devtools::install_github()`. Here, we will use the CRAN version.

```{r}
# Check if package installed. If not, install it.
if(!require('ciftiTools', quietly=TRUE)){
  install.packages('ciftiTools')
  # devtools::install_github('mandymejia/ciftiTools') # development version
}
```

Now we load the `ciftiTools` package. 

```{r}
library(ciftiTools)
```

Next, we indicate where to find the Connectome Workbench. This can be the full path to the Connectome Workbench executable file, or the path to its containing folder, in which case `ciftiTools` will locate the full path. Here, we will use the latter:

```{r}
# Replace '~/Desktop/workbench' with the actual path to 
#   the Connectome Workbench folder on your computer.
#   If successful, the path to the Workbench executable will be printed.
ciftiTools.setOption('wb_path', '~/Applications')

# For Tahoe
if (ciftiTools:::is_tahoe()) {
  options(ciftiTools.web_render = TRUE)
}
```

In this vignette, we will use example data included in the `ciftiTools` package. The files are originally from [NITRC](https://www.nitrc.org/frs/?group_id=454):

* The "MyelinAndCorrThickness" dtseries and dscalar files contain the same data: "MyelinMap_BC_decurv" and "corrThickness". (A dtseries file usually contains fMRI BOLD timeseries data, but that would be too large to include in the package.)
* The dlabel file contains three cortical parcellations
* The "ones" dscalar file is the only file to include subcortical voxels. All its data values are 1. 
* To reduce the size of `ciftiTools`, the dscalar and dlabel CIFTIs were resampled to 6k and the "ones" dscalar was resampled to 1k.

We will also use GIFTI files containing surface geometry.

```{r}
cifti_fnames <- ciftiTools.files()$cifti
surfL_fname <- ciftiTools.files()$surf["left"]
surfR_fname <- ciftiTools.files()$surf["right"]
```

These files are all in `fs_LR` coordinates, but `ciftiTools` can work with files created with other brain templates too, such as `fsaverage`. 

# Reading and Writing

### Reading

CIFTI files organize the gray matter of the brain into "grayordinates": vertices representing the left and right cortical surfaces, and voxels representing the subcortical gray matter structures and the cerebellum. A CIFTI file consists of two parts: (1) an XML header which contains all the metadata including medial wall locations, subcortical structure labels, and the subcortical volumetric mask; and (2) a NIFTI format matrix representing all the grayordinate data. These components are read in together with `read_cifti`:

```{r}
basename(cifti_fnames["dtseries"])
xii <- read_xifti(cifti_fnames["dtseries"])
```

The result of `read_cifti` is a `"xifti"` object. We can get an overview of it using its `summary` S3 method:

```{r}
xii # same as `summary(xii)`
```

By default, `read_cifti` will read-in all existing brainstructures. Set `brainstructures=c("left","right")`, for example, to read in just the left and right cortex data. The full set of choices for brainstructures is any combination of `"left"`, `"right"` and `"subcortical"`; `"all"` for all three; or, `"existing"` (default).

`"xifti"` objects are lists of three entries: `data` (the grayordinate data matrix, separated by brainstructure), `surf` (surface geometry), and `meta` (metadata, most of which is from the NIFTI XML header). `surf` distinguishes a `"xifti"` from a CIFTI file: the left and right cortical surface geometries are not included in CIFTI files, so they must be read from separate surface GIFTI files (ending in `surf.gii`). A surface must be compatible: the number of vertices must be the same, and each vertex in the CIFTI data must correspond to the vertex location in the corresponding GIFTI surface file. In this way, a `"xifti"` represents a combination of a CIFTI file with compatible GIFTI files for the cortical mesh.

We can add GIFTI surface geometries with `add_surf`:

```{r}
xii <- add_surf(xii, surfL=surfL_fname, surfR=surfR_fname)
xii
```

Alternatively, we could have provided the surface geometries at the outset of reading the CIFTI file:

```{r}
xii2 <- read_xifti(cifti_fnames["dtseries"], surfL_fname=surfL_fname, surfR_fname=surfR_fname)
all.equal(xii, xii2) # same result
```

To only read the CIFTI header, use `info_cifti`. Let's read the header of the dscalar file:

```{r}
xii_info <- ciftiTools::info_cifti(cifti_fnames["dscalar"])
str(xii_info, nchar.max=50) # shows header structure
```

To read in only certain columns of a CIFTI file, use the `idx` argument:

```{r}
read_xifti(cifti_fnames["dtseries"], idx=2) # second column only
```

### Writing

When a `"xifti"` object is written to files, the CIFTI components are placed in a CIFTI file and the surface geometries, if any, are placed in GIFTI files.

```{r}
out_dir <- tempdir()

write_xifti(
  xii, 
  file.path(out_dir, "my_cifti.dtseries.nii"), 
  file.path(out_dir, "my_L.surf.gii"), file.path(out_dir, "my_R.surf.gii")
)
```

With `separate_cifti`, a CIFTI can be separated and written into its component parts: the cortical data can be written to GIFTI metric or label files, and the subcortical data can be written to a NIFTI file. In addition, any ROIs or labels will also be written to files. The files are automatically named unless new file names are provided.

```{r}
# Use default names for everything except left cortex
separated_fnames = separate_cifti(
  cifti_fnames["dscalar_ones"], brainstructures="existing", 
  cortexL_fname="my_left_cortex.func.gii", write_dir = out_dir
)
# Files written to `out_dir`, or current working dir. if not specified
basename(separated_fnames)
```

Separated files can be read into R with the `oro.nifti`, `RNifti`, and `gifti` packages, and combined into a `"xifti"` object with `as.xifti`.

# Visualization

The cortical plots in `ciftiTools` are made possible by the `rgl` package. To prepare the R Markdown document for knitting we need to do the following: 

```{r}
library(rgl)

if (ciftiTools:::is_tahoe()) {
  chromote::local_chrome_version("latest-stable", binary = "chrome-headless-shell")
}

rgl::setupKnitr()

# Sometimes the first OpenGL window does not render properly.
rgl::open3d(); rgl::close3d()
```

Now let's take a look! 

### Surface visualization

`view_xifti_surface(xii)` displays the cortical data on the surface mesh. This function has several primary arguments:

* `color_mode` specifies the nature of the data values: `"sequential"`, `"qualitative"` and `"diverging"`. If it is not provided, a default mode that makes sense for the data will be used.
* `colors` specifies the color palette to use. If it is not provided, a default palette that makes sense for the `color_mode` is used.
* `idx` controls which column(s) to display.
* `widget` and `fname` control the output type. If `fname` is not provided, an interactive plot is created: by default, an OpenGL window if the length of `idx` is one, and an embedded HTML widget if the length of `idx` is greater than one. `widget` can be used to override this default. On the other hand, if `fname` is provided, static image files (png) for each `idx` are created, unless `fname` ends in `.html` in which case an interactive html file will be saved. Lastly, although OpenGL windows can be embedded in R Markdown documents for knitting through screenshotting capabilities, the results are unstable. We recommend embedding either widgets or static images (using `fname` to save temporary image files, and then adding them with `:knitr::include_graphics`). Refer to the source code of this vignette to see how this works.
* `surfL` and `surfR` specify the surface geometry to plot the data on. If not provided, the surfaces in the `"xifti"` object is used, if available; if not, the "inflated" surfaces included in `ciftiTools` are used.

Let's see a couple examples. We'll first look at two indices from the dtseries file. Notice how a slider is added to control which index is being visualized. Also note that the default surfaces are automatically used for visualization. 
(In our case, the included surfaces _are_ the default surfaces.) Try clicking and dragging around to rotate, and scrolling to zoom in and out. 

```{r, fig.cap="dtseries file; first column; sequential palette", fig.width=5, fig.height=4.2}
view_xifti_surface(xii, idx=seq(2), zlim=c(1,2), title='color_mode = "sequential"')
```

Next we'll look at the single-column dlabel file. 

```{r, fig.cap="dlabel file; first label; palette from label metadata", fig.width=5}
view_xifti_surface(
  read_cifti(cifti_fnames["dlabel"]), 
  # Interactively, a color legend that displays the label names will also be printed.
  legend_ncol=5, 
  title='color_mode = "qualitative"'
)
```

### Volume visualization

`view_xifti_volume(xii)` displays the subcortical data in slices. To view interactively in an interactive session, set `interactive=TRUE`. By default, a series of slices is displayed overlaid on the MNI template. The orientation and numbers of slices can be adjusted. A .png or .pdf file can be written with `fname`. The same color arguments from `view_xifti_surface` work too: `colors`, `color_mode` and `zlim`.

```{r, fig.cap="Subcortical data (all ones)", fig.height=6, fig.width=4}
# cifti_fnames["dscalar_ones"] is the only file with subcortical data
xii <- read_cifti(cifti_fnames["dscalar_ones"], brainstructures="subcortical")
view_xifti_volume(xii)
```

```{r eval=FALSE}
# For information only, since papaya viewer cannot be embedded in knitted file
view_xifti_volume(xii, interactive = TRUE)
```

The S3 method `plot(xii)` will display all the data present in the `xifti` using `view_xifti_surface`, `view_xifti_volume`, or both. If both plots are made, the color mode and range will be shared between the two plots. 

### Tips and tricks

`NA` values will be uncolored. You can use this to, for example, only color values meeting a certain threshold.

```{r, fig.cap="dscalar; using NA values to selectively color", fig.width=5}
xii <- read_cifti(cifti_fnames["dscalar"], resamp_res=12000)
# Convert to z-scores
xii <- scale_xifti(xii)
# Threshold
xii <- transform_xifti(xii, function(x){ifelse(x<2, NA, x)})
view_xifti_surface(
  xii, title='MyelinMap_BC_decurv: z > 2', 
  zlim=c(2,3), NA_color="#505560"
)
```

Something to note is that the shadows from 3D rendering will darken colors in some folded regions of the brain. Another consideration is that the colors are smoothed continuously. In some situations, these default 3D shading properties may be undesirable. The `shadows` and `material` arguments can be used to adjust aspects of 3D shading.

To demonstrate, we'll first remove shadows, and then remove both shadows and color smoothing. We'll also use static images here to show how that works.

```{r, fig.cap="MyelinMap_BC_decurv: z > 2, with different lighting options", fig.show="hold"}
xii_plts <- c(tempfile(fileext=".png"), tempfile(fileext=".png"), tempfile(fileext=".png"))

view_xifti_surface(
  xii, hemisphere="left", title='Default',
  zlim=c(2,3), NA_color="#505560", fname=xii_plts[1]
)

view_xifti_surface(
  xii, hemisphere="left", title='Less shadows',
  zlim=c(2,3), NA_color="#505560", fname=xii_plts[2],
  shadows=0
)

view_xifti_surface(
  xii, hemisphere="left", title='Less shadows & exact colors', 
  zlim=c(2,3), NA_color="#505560", fname=xii_plts[3],
  material=list(lit=FALSE, smooth=FALSE)
)

knitr::include_graphics(xii_plts, dpi=250)
```

# Resampling and smoothing

### Resampling

`ciftiTools` can resample CIFTI files to a lower resolution. Here, we resample the 32k dtseries file to 2k. (The number refers to the count of vertices on a single hemisphere.) We also provide the surfaces and resample them in conjunction.

```{r}
resampled_xii_fname <- "my_new_resampled.dtseries.nii"
resampled_surfL_fname <- "my_resampled_surfL.surf.gii"
resampled_surfR_fname <- "my_resampled_surfR.surf.gii"
  
xii_2k <- resample_cifti(
  cifti_fnames["dtseries"], resampled_xii_fname,
  resamp_res = 2000,
  surfL_fname, surfR_fname,
  resampled_surfL_fname, resampled_surfR_fname,
  write_dir = out_dir
)
basename(xii_2k)
```

Resampling can also be performed while reading a CIFTI file into R. 

```{r}
read_cifti(cifti_fnames["dscalar"], resamp_res=2000)
```

### Smoothing

Use `smooth_cifti` to perform smoothing. Like resampling, this function works on both CIFTI files and `"xifti"` objects.

```{r}
smoothed_xii_fname <- "my_smoothed_cifti.dtseries.nii"

# Smoothing a CIFTI file
smooth_cifti(
  cifti_fnames["dtseries"], file.path(out_dir, smoothed_xii_fname),
  surf_FWHM=2, vol_FWHM=2,
  surfL_fname=surfL_fname, surfR_fname=surfR_fname,
  subcortical_zeroes_as_NA=TRUE
)
```

We'll plot this smoothed CIFTI file with a diverging palette. Let's try both the default pallete and an RColorBrewer palette. 

```{r, fig.cap="Smoothed CIFTI, default palette", fig.width=5}
view_xifti_surface(
  read_cifti(file.path(out_dir, smoothed_xii_fname)), 
  surfL=surfL_fname, surfR=surfR_fname, 
  zlim=c(.6,1.8), color_mode="diverging", title="Default diverging palette"
)
```

```{r, fig.cap="Smoothed CIFTI, RColorBrewer palette", fig.width=5}
view_xifti_surface(
  read_cifti(file.path(out_dir, smoothed_xii_fname)), 
  surfL=surfL_fname, surfR=surfR_fname, 
  zlim=c(.6,1.8), color_mode="diverging", colors="Spectral",
  title="RColorBrewer diverging palette"
)
```

# Manipulation & Math

You can treat the `"xifti"` as a data matrix with the base R functions `as.matrix`, `nrow`, and `ncol`.

```{r}
xiiL <- read_xifti(cifti_fnames["dscalar"], brainstructures="left")
dim(as.matrix(xiiL))
```

`apply_xifti` applies a function along the rows or columns of a `"xifti"`. The base R `apply` function also works on `"xifti"` objects. The difference is that when applying a function along the rows, the former will return a `"xifti"` whereas the latter will return a data matrix. For example, we can compute the mean at each data location like so:

```{r}
gmeans <- apply_xifti(xiiL, 1, mean)
gmeans
```

And we can compute quantiles of each measurement like so:

```{r}
cquants <- apply_xifti(xiiL, 2, quantile, c(.1, .5)) # Quantiles of each column
cquants
```

`combine_xifti` combines multiple `"xifti"`s with different brainstructures.

```{r}
xiiR <- read_xifti(cifti_fnames["dscalar"], brainstructures="right")
xii <- combine_xifti(xiiL, xiiR)
xii
```

`convert_xifti` converts the intent of a `"xifti"`.

```{r}
convert_xifti(xii, "dtseries") # Convert from dscalar to dtseries (don't use it)
```

`merge_xifti` concatenates data matrices from multiple `"xifti"`s column-wise.

```{r}
xii <- merge_xifti(xii, xii) # Columns are repeated twice now.
```

`remove_xifti` removes a brainstructure(s) from a `"xifti"`.

```{r}
xii <- remove_xifti(xii, "cortex_left") # Now only the right cortex data is included.
xii
```

`select_xifti` subsets or re-orders the columns of the data matrix of a `"xifti"`.

```{r}
xii$meta$cifti$names <- paste("Column", seq(4))
xii <- select_xifti(xii, c(4,3,2)) # Reverse column order & drop the first.
xii
```

S3 methods allow for univariate transformations of `"xifti"` objects as well as arithmetic operations of multiple `"xifti"` objects.

```{r}
max(as.matrix(xii))
xii <- 1 - exp(xii) / (xii * 2 + 3)
max(as.matrix(xii))
```

# Working with surfaces

`ciftiTools` also includes functionality for working with surface geometry GIFTI files separately from any data. Surfaces that are read in are `"surf"` objects:

```{r}
# Reading
surf <- read_surf(surfL_fname)
surf
```

These can be written back to GIFTI files, visualized, and resampled. Resampling can be performed on the `"surf"` objects or the surface GIFTI files directly:


```{r, fig.width=4}
# Writing
write_surf_gifti(surf, file.path(out_dir, "my.L.surf"))

# Resampling and visualizing
surf <- resample_surf(surf, resamp_res=2000, "left")
plot(surf)
```

```{r}
# Resample a GIFTI file
resample_gifti(surfL_fname, file.path(out_dir, "my.L.2k.surf.gii"), "left", resamp_res=8000)
```

As mentioned earlier, the inflated surface is included as a GIFTI file and can be accessed with `ciftiTools.files()`. It is used as default for visualizing data with `view_xifti_surface`. But two more surfaces are available through the function `load_surf`:

```{r, fig.cap="Two alternative surfaces"}
xii <- as.xifti(
  surfL = load_surf("left", "very inflated"),
  surfR = load_surf("right", "midthickness")
)
xii_fname <- tempfile(fileext=".png")
plot(xii, title = "Left very inflated | Right midthickness", fname=xii_fname)
knitr::include_graphics(xii_fname, dpi=180)
```

Lastly, let's demonstrate the ability to plot vertices and edges. (This can also be done when plotting `"xifti"` objects with data, using the same arguments.)

```{r, fig.cap="Surface with vertices drawn in blue", fig.width=3}
# Recall that surf was resampled to 2k
plot(surf, vertex_size=3, view="lateral", vertex_color="blue", zoom=1)
```

```{r, fig.cap="Surface with edges drawn in black", fig.width=3}
plot(surf, edge_color="black", view="lateral", zoom=1)
```

### Mesh operations

`ciftiTools` includes several functions for operations that depend on the vertex adjacencies of the surface mesh.  

`edit_mask_surf` can erode, dilate, or extract the border vertices of a binary mask. (Here we'll also demonstrate how to use `view_comp` to composite multiple static plots into one image.)

```{r fig.height=4.2, fig.width=5.5, fig.cap="Dilating a mask"}
xii <- load_parc("Schaefer_400")
xii <- remove_xifti(xii, "cortex_right")

# Select this parcel: 17networks_LH_DefaultA_PCC_1
parc_name <- rownames(xii$meta$cifti$labels[[1]])[5]
parc_name2 <- gsub("17networks_LH_", "", parc_name)

# Create a new `"xifti"` with 1 at that parcel, 0 elsewhere.
xii <- apply_xifti(xii, 1, function(x){ifelse(x==4, 1, 0)})

# Create another, dilating that parcel mask by 3 layers.
xii_d3 <- edit_mask_surf(
  xii$data$cortex_left[,1], mwall=rep(TRUE, 32492), 
  surf=load_surf("left"),
  do="dilate", depth=3
)

# Merge the two `"xifti"` objects and convert back to dlabel.
xii <- newdata_xifti(xii, cbind(as.matrix(xii), xii_d3))
xii <- convert_xifti(xii, "dlabel", colors="blue", labels=c("(masked)", parc_name2))

# Plot together. 
# (Use fnames[3] for the legend and fnames[4] for the composite image.)
fnames <- paste0(c(tempfile(), tempfile(), tempfile(), tempfile()), ".png")
plot(xii, title="Original", idx=1, fname=fnames[1], legend_fname=fnames[3])
plot(xii, title="Dilated", idx=2, fname=fnames[2], legend_fname=fnames[3])
view_comp(fnames[seq(2)], legend=fnames[3], fname=fnames[4], nrow=1)
knitr::include_graphics(fnames[4], dpi=180)
```

`even_vert_samp` obtains a subset of vertices sampled evenly across the mesh. It works by downsampling the mesh and then identifying the original vertices closest to the vertices on the downsampled mesh. 

```{r, fig.cap="500 evenly-sampled vertices"}
# Load a left-hemisphere surface.
surfL <- load_surf("left")

# Get 500 vertices evenly sampled on the mesh.
v <- even_vert_samp(surfL, 500)

# Get a mask of the 32k vertices indicating which are in the sample.
v2 <- seq(nrow(surfL$vertices)) %in% v

# Make a xifti having "1" for in-mask vertices, "0" elsewhere.
xii <- as.xifti(as.matrix(v2*1), surfL=surfL)

xii <- convert_xifti(xii, "dlabel", colors="black")
xii <- resample_cifti(xii, resamp_res=50000) # for viz.
xii_plt <- tempfile(fileext=".png")
plot(xii, fname=xii_plt)
knitr::include_graphics(xii_plt, dpi=180)
```

# Working with parcellations

### Loading included parcellations 

`ciftiTools` includes several commonly-used parcellations which can be loaded in as a single-column dlabel `"xifti"` with `load_parc`: the 2018 Schaefer parcellations at 100, 400, and 1000 parcels resolutions, and the 2011 Yeo 7- and 17-network parcellations. Refer to `?load_parc` for their corresponding references. Any other parcellations can be read in with `read_cifti`. 

```{r fig.cap="Schaefer 400 parcellation (label colors)", fig.width=5}
parc <- load_parc("Schaefer_400")
# parc <- read_cifti("my_parcellation.dlabel.nii")

view_xifti_surface(parc, title="Schaefer 400")
```

`parc_add_subcortex` will add new label levels and "parcels" data for each subcortical structure in the MNI template. 

```{r}
# Before `parc_add_subcortex`
max(parc) # 400 parcels
summary(parc)
```

```{r}
parc <- parc_add_subcortex(parc)
max(parc) # 400 cortex parcels + 19 subcortex parcels
summary(parc) # After `parc_add_subcortex`
```

### Applying a parcellation to data

`apply_parc` will apply a function to one `"xifti"` across each parcel specified by another `"xifti"`. A common example is computing the average timeseries of each parcel:

```{r}
parc <- load_parc("Yeo_7")
parc_tab <- parc$meta$cifti$labels[[1]]
rbind(head(parc_tab), tail(parc_tab)) # Keys 0-51 (52 total)
```

```{r}
dat <- read_cifti(ciftiTools.files()$cifti["dtseries"])
dim(dat)
```

```{r}
pmean <- apply_parc(
  dat, parc, 
  FUN=colMeans, # get average timeseries
  na.rm=TRUE # medial wall vals in `dat` will be NA
)
dim(pmean) # 52 parcels x 2 data columns
```

For real timeseries data, `cor(t(pmean))` would give the parcels x parcels functional connectivity matrix.

`apply_parc` also has an argument which converts the result to a `"xifti"` object. Each locations' value will be the value of its corresponding parcel.

```{r, fig.cap="Median values for the 51 components in the Yeo 7 parcellation"}

# Obtain the min/max, quartiles, and mean value
#   of the second data column, within each parcel.
psumm <- apply_parc(
  select_xifti(dat, idx=2),
  parc, mwall_value=0, 
  FUN=summary,
  return_as="xifti"
)

plt_fname <- tempfile(fileext=".png")
plot(psumm, idx=3, title=psumm$meta$cifti$names[3], 
     colors="magma", zlim=c(1, 3), fname=plt_fname)
knitr::include_graphics(plt_fname, dpi=180)
```

# Citing `ciftiTools`

A citation for the package itself can be printed with:

```{r}
citation("ciftiTools")
```

Refer to the README for citation information for the surfaces, parcellations, and other data included in `ciftiTools`, as well as the Connectome Workbench. Also check the DESCRIPTION file to get a list of R packages used, including `rgl` and `papayar`. 
