---
title: "`ciftiTools` Mini Demo"
author: "Damon Pham & Amanda Mejia"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
  toc: true
  keep_md: true
vignette: >
  %\VignetteIndexEntry{`ciftiTools` Mini Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
library(knitr)
knitr::opts_chunk$set(autodep = TRUE, cache = FALSE)
```

`ciftiTools` is an R package for working with CIFTI-2 format brain imaging data. Used in conjunction with GIFTI surface geometry files, CIFTI files enable surface-based analysis of gray matter data, which has several advantages over traditional volumetric/voxel-based analysis. Because of this, the CIFTI-2 format is used by recent neuroimaging studies including the Human Connectome Project (HCP). `ciftiTools` supports reading, writing, visualizing, resampling, and other operations for CIFTI files with the `".dscalar.nii"`, `".dtseries.nii"`, and `".dlabel.nii"` intents. Several of these operations are made possible by the [Connectome Workbench](https://www.humanconnectome.org/software/connectome-workbench). 

_Several key operations in `ciftiTools`, including reading and writing CIFTI files, are made possible by the [Connectome Workbench](https://www.humanconnectome.org/software/connectome-workbench). Since the Workbench must be locally installed, this vignette made for CRAN can only demonstrate a few functions. Please refer to the [full vignette on GitHub](https://htmlpreview.github.io/?https://github.com/mandymejia/ciftiTools/blob/master/vignettes/ciftiTools_vignette.html)._

To get started, the first time you use `ciftiTools`, install it from either CRAN with `install.packages()` or Github with `devtools::install_github()`. Here, we will use the CRAN version.

```{r}
# Check if package installed. If not, install it.
if(!require('ciftiTools', quietly=TRUE)){
  install.packages('ciftiTools')
  # devtools::install_github('mandymejia/ciftiTools') # development version
}
```

Now we load the `ciftiTools` package. 

```{r}
library(ciftiTools)
```


Next, we indicate where to find the Connectome Workbench. This can be the full path to the Connectome Workbench executable file, or the path to its containing folder, in which case `ciftiTools` will locate the full path. Here, we will use the latter. 

_Note that this step is skipped with `eval=FALSE` for the purpose of including the vignette on CRAN._

```{r eval=FALSE}
# Replace '~/Applications' with the actual path to the 
#   Connectome Workbench folder on your computer. If
#   successful, the Workbench executable path will be printed.
ciftiTools.setOption('wb_path', '~/Applications')
```

Since reading in CIFTI files requires the Connectome Workbench, let's instead use `load_parc` to load one of the cortex parcellations included in `ciftiTools`.

```{r}
xii <- load_parc() # Loads the Schaefer 100 parcellation.
xii # Summary of the `"xifti"` object.
```

# Visualization

Cortex plots in `ciftiTools` are made possible by the `rgl` package. To prepare the R Markdown document for knitting we need to do the following: 

```{r}
library(rgl)
rgl::setupKnitr()

# Sometimes the first OpenGL window does not render properly.
rgl::open3d(); rgl::close3d()

# These are also required.
library(manipulateWidget)
library(ggpubr)
```

Now we can use `view_xifti_surface(xii)` to display the cortical data on the surface mesh. This function has several primary arguments:

* `color_mode` specifies the nature of the data values: `"sequential"`, `"qualitative"` and `"diverging"`. If it is not provided, a default mode that makes sense for the data will be used.
* `colors` specifies the color palette to use. If it is not provided, a default palette that makes sense for the `color_mode` is used.
* `idx` controls which column(s) to display.
* `widget` and `fname` control the output type. If `fname` is not provided, an interactive plot is created: by default, an OpenGL window if the length of `idx` is one, and an embedded HTML widget if the length of `idx` is greater than one. `widget` can be used to override this default. On the other hand, if `fname` is provided, static image files (png) for each `idx` are created, unless `fname` ends in `.html` in which case an interactive html file will be saved. Lastly, both OpenGL windows and HTML widgets can be embedded in R Markdown documents for knitting; refer to the source code of this vignette to see how this works.
* `surfL` and `surfR` specify the surface geometry to plot the data on. If not provided, the surfaces in the `"xifti"` object is used. But if those are also unavailable, the "inflated" surfaces included in `ciftiTools` are used.

Let's plot our `"xifti"` object. Note that interactively, a color legend which displays the label names will also be printed.

```{r, fig.cap="Schaefer 100 parcellation", rgl=TRUE, format="jpg", fig.height=2.1, fig.width=2.5}
# Normally `cex.title` doesn't need to be set, as it defaults to a good choice.
#   But when knitting static images this way, the default becomes a bit too big
#   based on how knitting works.
view_xifti_surface(xii, idx=1, title='Schaefer 100', cex.title=1.2)
```

# Manipulation

`add_surf` adds surface geometry to the `"xifti"` object.

```{r}
xii <- add_surf(xii, "midthickness", "midthickness")
xii
```

`remove_xifti` removes a brain structure.

```{r}
xii <- remove_xifti(xii, c("cortex_right", "surf_right"))
xii
```

`transform_xifti` applies a function to the data values. Let's isolate and view the frontal pole cortex. Note the midthickness surface that was added will be used now, instead of the default inflated surface.

```{r, fig.cap="Plotting the FPole parcel", rgl=TRUE, format="jpg", fig.height=2, fig.width=1.3}
label_to_viz <- "17networks_LH_DefaultB_FPole_1"
key_idx <- which(rownames(xii$meta$cifti$labels$parcels)==label_to_viz)
key <- xii$meta$cifti$labels$parcels$Key[key_idx]
xii <- transform_xifti(xii, function(v){ifelse(v==key, v, 0)})
view_xifti_surface(xii)
```

# Citing `ciftiTools`

A citation for the package itself can be printed with:

```{r}
citation("ciftiTools")
```

Refer to the README for citation information for the surfaces, parcellations, and other data included in `ciftiTools`, as well as the Connectome Workbench. Also check the DESCRIPTION file to get a list of R packages used, including `rgl` and `papayar`. Lastly, check out the [full vignette on GitHub](https://htmlpreview.github.io/?https://github.com/mandymejia/ciftiTools/blob/master/vignettes/ciftiTools_vignette.html).
