---
title: "`ciftiTools` Demo"
author: "Damon Pham & Amanda Mejia"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
  toc: true
  keep_md: true
vignette: >
  %\VignetteIndexEntry{`ciftiTools` Demo}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r include=FALSE}
library(knitr)
knitr::opts_chunk$set(autodep = TRUE, cache = FALSE)

is_knit <- knitr::is_html_output() || knitr::is_latex_output()
#if (is_knit) { options(rgl.useNULL = TRUE) }

knit_pic <- function(dpi=200){
  pic <- tempfile(fileext=".png")
  rgl:::rgl.snapshot(pic)
  rgl::close3d()
  #suppressMessages(rgl::snapshot3d(pic, webshot=FALSE))
  knitr::include_graphics(pic, dpi=dpi)
}
```

`ciftiTools` is an R package for working with CIFTI-2 format brain imaging data. Used in conjunction with GIFTI surface geometry files, CIFTI files enable surface-based analysis of gray matter data, which has several advantages over traditional volumetric/voxel-based analysis. Because of this, the CIFTI-2 format is used by recent neuroimaging studies including the Human Connectome Project (HCP). `ciftiTools` supports reading, writing, visualizing, resampling, and other operations for CIFTI files with the `".dscalar.nii"`, `".dtseries.nii"`, and `".dlabel.nii"` intents. Several of these operations are made possible by the [Connectome Workbench](https://www.humanconnectome.org/software/connectome-workbench). 

To get started, the first time you use `ciftiTools`, install it from either CRAN with `install.packages()` or Github with `devtools::install_github()`. Here, we will use the CRAN version.

```{r}
# Check if package installed. If not, install it.
if(!require('ciftiTools', quietly=TRUE)){
  install.packages('ciftiTools')
  # devtools::install_github('mandymejia/ciftiTools') # development version
}
```

Now we load the `ciftiTools` package. 

```{r}
library(ciftiTools)
```

Next, we indicate where to find the Connectome Workbench. This can be the full path to the Connectome Workbench executable file, or the path to its containing folder, in which case `ciftiTools` will locate the full path. Here, we will use the latter:

```{r}
# Replace '~/Desktop/workbench' with the actual path to 
#   the Connectome Workbench folder on your computer.
#   If successful, the path to the Workbench executable will be printed.
ciftiTools.setOption('wb_path', '~/Applications')
```

In this vignette, we will use example data included in the `ciftiTools` package. The files are originally from [NITRC](https://www.nitrc.org/frs/?group_id=454):

* The "MyelinAndCorrThickness" dtseries and dscalar files contain the same data: "MyelinMap_BC_decurv" and "corrThickness". (A dtseries file usually contains fMRI BOLD timeseries data, but that would be too large to include in the package.)
* The dlabel file contains three cortical parcellations
* The "ones" dscalar file is the only file to include subcortical voxels. All its data values are 1. 
* To reduce the size of `ciftiTools`, the dscalar and dlabel CIFTIs were resampled to 6k and the "ones" dscalar was resampled to 1k.

We will also use GIFTI files containing inflated surface geometry.

```{r}
cifti_fnames <- ciftiTools.files()$cifti
surfL_fname <- ciftiTools.files()$surf["left"]
surfR_fname <- ciftiTools.files()$surf["right"]
```

# Reading and Writing

### Reading

CIFTI files organize the gray matter of the brain into "grayordinates": vertices representing the left and right cortical surfaces, and voxels representing the subcortical gray matter structures and the cerebellum. A CIFTI file consists of two parts: (1) an XML header which contains all the metadata including medial wall locations, subcortical structure labels, and the subcortical volumetric mask; and (2) a NIFTI format matrix representing all the grayordinate data. These components are read in together with `read_cifti`:

```{r}
basename(cifti_fnames["dtseries"])
xii <- read_xifti(cifti_fnames["dtseries"])
```

The result of `read_cifti` is a `"xifti"` object. We can get an overview of it using its `summary` S3 method:

```{r}
xii # same as `summary(xii)`
```

By default, `read_cifti` will read-in all existing brainstructures. Set `brainstructures=c("left","right")`, for example, to read in just the left and right cortex data. The full set of choices for brainstructures is any combination of `"left"`, `"right"` and `"subcortical"`, `"all"` for all three, or `"existing"` (default).

`"xifti"` objects are lists with entries `data` (the grayordinate data matrix, separated by brainstructure), `surf` (surface geometry), and `meta` (metadata, most of which is from the NIFTI XML header). `surf` distinguishes a `"xifti"` from a CIFTI file: the left and right cortical surface geometries are not included in CIFTI files, so they must be read from separate surface GIFTI files (ending in `surf.gii`). The surface must be compatible: the number of vertices must be the same, and each vertex in the CIFTI data must correspond to the vertex location in the corresponding GIFTI surface file. In this way, a `"xifti"` represents a combination of a CIFTI file with compatible GIFTI files for the cortical mesh.

We can add GIFTI surface geometries with `add_surf`:

```{r}
xii <- add_surf(xii, surfL=surfL_fname, surfR=surfR_fname)
xii
```

Alternatively, we could have provided the surface geometries at the outset of reading the CIFTI file:

```{r}
xii2 <- read_xifti(cifti_fnames["dtseries"], surfL_fname=surfL_fname, surfR_fname=surfR_fname)
all.equal(xii, xii2) # same result
```

To only read the CIFTI header, use `info_cifti`. Let's read the header of the dscalar file:

```{r}
xii_info <- ciftiTools::info_cifti(cifti_fnames["dscalar"])
str(xii_info, nchar.max=50) # shows header structure
```

To read in only certain columns of a CIFTI file, use the `idx` argument:

```{r}
read_xifti(cifti_fnames["dtseries"], idx=2) # second column only
```

### Writing

When a `"xifti"` object is written to files, the CIFTI components are placed in a CIFTI file and the surface geometries, if any, are placed in GIFTI files.

```{r}
out_dir <- tempdir()

write_xifti(
  xii, 
  file.path(out_dir, "my_cifti.dtseries.nii"), 
  file.path(out_dir, "my_L.surf.gii"), file.path(out_dir, "my_R.surf.gii")
)
```

With `separate_cifti`, a CIFTI can be separated and written into its component parts: the cortical data can be written to GIFTI metric or label files, and the subcortical data can be written to a NIFTI file. In addition, any ROIs or labels will also be written to files. The files are automatically named unless a new file name is provided.

```{r}
# Use default names for everything except left cortex
separated_fnames = separate_cifti(
  cifti_fnames["dscalar_ones"], brainstructures="existing", 
  cortexL_fname="my_left_cortex.func.gii", write_dir = out_dir
)
# Files written to `out_dir`, or current working dir. if not specified
basename(separated_fnames)
```

Separated files can be read into R with the `oro.nifti`, `RNifti`, and `gifti` packages, and combined into a `"xifti"` object with `as.xifti`.

# Visualization

The cortical plots in `ciftiTools` are made possible by the `rgl` package. To prepare the R Markdown document for knitting we need to do the following: 

```{r}
library(rgl)
rgl::setupKnitr()

# Sometimes the first OpenGL window does not render properly.
rgl::open3d(); rgl::close3d()

# These are also required.
library(manipulateWidget)
library(ggpubr)
```

Now let's take a look! 

### Surface visualization

`view_xifti_surface(xii)` displays the cortical data on the surface mesh. This function has several primary arguments:

* `color_mode` specifies the nature of the data values: `"sequential"`, `"qualitative"` and `"diverging"`. If it is not provided, a default mode that makes sense for the data will be used.
* `colors` specifies the color palette to use. If it is not provided, a default palette that makes sense for the `color_mode` is used.
* `idx` controls which column(s) to display.
* `widget` and `fname` control the output type. If `fname` is not provided, an interactive plot is created: by default, an OpenGL window if the length of `idx` is one, and an embedded HTML widget if the length of `idx` is greater than one. `widget` can be used to override this default. On the other hand, if `fname` is provided, static image files (png) for each `idx` are created, unless `fname` ends in `.html` in which case an interactive html file will be saved. Lastly, both OpenGL windows and HTML widgets can be embedded in R Markdown documents for knitting; refer to the source code of this vignette to see how this works.
* `surfL` and `surfR` specify the surface geometry to plot the data on. If not provided, the surfaces in the `"xifti"` object is used. But if those are also unavailable, the "inflated" surfaces included in `ciftiTools` are used.

Let's see an example using each `color_mode` option. Note how the included surfaces are used in the first plot, but if none are present as in the second and third plots, the default surfaces are automatically used for visualization. (In our case, the included surfaces _are_ the default surfaces.) We'll also make the second plot interactive by requesting display of two `idx`. Try clicking and dragging around the second plot to rotate, and scrolling to zoom in and out. Note that the first and third plots are OpenGL window snapshots, and that the second plot is an embedded HTML widget.

```{r, fig.cap="dtseries file; first column; sequential palette"}
view_xifti_surface(xii, idx=1, zlim=c(1,2), title='color_mode = "sequential"')
knit_pic()
```

```{r fig.height=4.2, fig.width=5.5}
xii <- read_cifti(cifti_fnames["dscalar"]) # no GIFTI included, so the default inflated surface is used.
view_xifti_surface(
  xii, idx=1:2, zlim=c(0,5), color_mode = "diverging",
  title='color_mode = "diverging"', cex.title=1.3
)
```

```{r, fig.cap="dlabel file; first label; palette from label metadata"}
view_xifti_surface(
  read_cifti(cifti_fnames["dlabel"]), 
  # Interactively, a color legend that displays the label names will also be printed.
  legend_ncol=5, 
  title='color_mode = "qualitative"'
)
knit_pic()
```
